---
title: "Outlier Analysis"
subtitle: "Supplement to Praet et al. 2019, Turbidite stratigraphy in proglacial lakes; deciphering turbidite trigger mechanisms using a statistical approach"
output:
  md_document:
    variant: markdown_github
  pdf_document: default
  html_document: default
---

In order to identify varves that potentially include a turbidite, we performed an outlier study on varve thickness records. Our working hypothesis is that a varve containing an extra turbidite (or a slump deposit) would be characterized by an anomalous varve thickness – an outlier – in comparison with the steady-state, or background bulk varve thickness distribution.

# Dependencies

First we load the required R-packages. For the outlier analysis the `extremevalues` R-package by van der Loo (2010) was used.

```{r message=FALSE, warning=FALSE}
library(extremevalues)
library(dplyr)
library(knitr)
library(kableExtra)
```

# Outlier Analysis

In order to objectively identify the outliers in varve thickness, the dataset was analysed with the “outlier detection method I”. First, a lognormal distribution model was fit to the data falling below the 0.95 or 0.75 quantile (set by `Flim`). An outlier varve was then defined as having a thickness value within the tail of the distribution where, according to the fitted model, less than 1 observation (set by `rho`) is expected to occur given the original number of data points. Hence, a varve is identified as an outlier when its thickness is larger than what can be expected based on 95 % or 75 % of the data, respectively. Plots of the thickness histogram, fitted distribution and QQ-plots are generated.

```{r echo=TRUE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
for(i in list.files("data")[grep("*.csv", list.files())]){
        name = tools::file_path_sans_ext(i)
        # read in the thickenss files
        data_set <- read.csv(i)[,c(1,2)]
        y <- data_set$thickness
        y <- sort(y[!is.na(y)])
        # detect outliers usign the getOutliers function. rho=c(1,1)
        L_0_75 = getOutliers(y,
                             method = "I",
                             distribution= "lognormal",
                             FLim=c(0,0.75),
                             rho = c(0.01,1))
        
        L_0_95 = getOutliers(y,
                             method = "I",
                             distribution= "lognormal",
                             FLim=c(0,0.95),
                             rho = c(0.01,1))

        # create outputs for 0_75
        dir.create("output_methodI_0_75", showWarnings = FALSE)
        
        ## write data to a file
        df <- do.call(rbind,lapply(L_0_75, data.frame, stringsAsFactors=FALSE))
        names(df) = paste(name,"methodI_0_75",sep="_")
        df %>% kable() %>% print() %>% cat()
        # cat("\n")
        write.csv(df, file.path("output_methodI_0_75",
                                paste(strsplit(i,".csv")[[1]],
                                      "_statistics.csv",sep = "")), row.names = T)
        
        ## generate histogram
        hist(y, probability = T, breaks = 100, col = "grey", main = i)
        curve(dlnorm(x, meanlog = L_0_75$mu,  sdlog = L_0_75$sigma),
              col="darkblue", lwd=2, add = T)
        ## save to file
        pdf(file.path("output_methodI_0_75",
                       paste(strsplit(i,".csv")[[1]],"_hist.pdf",sep = "")))
        hist(y, probability = T, breaks = 100, col = "grey", main = i)
        curve(dlnorm(x, meanlog = L_0_75$mu,  sdlog = L_0_75$sigma),
              col="darkblue", lwd=2, add = T)
        dev.off()
        
        ## generate qqplot
        qqFitPlot(y, L_0_75)
        ## save to file
        pdf(file.path("output_methodI_0_75",
                     paste(strsplit(i,".csv")[[1]],"_QQ.pdf",sep = "")))
        qqFitPlot(y, L_0_75)
        dev.off()
        
        # create outputs for 0_95
        dir.create("output_methodI_0_95", showWarnings = FALSE)
        
        ## print data and write to a file
        df <- do.call(rbind,
                      lapply(L_0_95,data.frame, stringsAsFactors=FALSE))
        names(df) = paste(name,"methodI_0_95",sep="_")
        df %>% kable() %>% print() %>% cat()
        # cat("\n")
        write.csv(df, file.path("output_methodI_0_95",
                                paste(strsplit(i,".csv")[[1]],
                                      "_statistics.csv",
                                      sep = "")), row.names = T)
        
        ## generate histogram
        hist(y, probability = T, breaks = 100, col = "grey", main = i)
        curve(dlnorm(x, meanlog = L_0_95$mu,  sdlog = L_0_95$sigma),
              col="darkblue", lwd=2, add = T)
        ## save to file
        pdf(file.path("output_methodI_0_95",
                       paste(strsplit(i,".csv")[[1]],"_hist.pdf",sep = "")))
        hist(y, probability = T, breaks = 100, col = "grey", main = i)
        curve(dlnorm(x, meanlog = L_0_95$mu,  sdlog = L_0_95$sigma),
              col="darkblue", lwd=2, add = T)
        dev.off()
        
        ## generate qqplot
        qqFitPlot(y, L_0_95)
        ## save to file
        pdf(file.path("output_methodI_0_95",
                       paste(strsplit(i,".csv")[[1]],"_QQ.pdf",sep = "")))
        qqFitPlot(y, L_0_95)
        dev.off()
        

} 
```

